---
documentclass: article
fontsize: 12pt
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2: 
    toc: false
    latex_engine: xelatex
    includes:
      in_header: preamble.sty
      before_body: titlepage.sty
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven) # to read SAS files
library(rprojroot)
knitr::opts_chunk$set(echo = TRUE)
```
# Abstract {-}

# Introduction and data exploration
```{r load, include=FALSE, message=FALSE, warning=FALSE}
data <- read_sas(find_root_file("data/hearing500lr.sas7bdat", 
     criterion = has_file("LDA.Rproj"))) %>%
  mutate(side = as.factor(side),
         id = as.factor(id),
         age_start = age - TIME)
summary_data <- data %>%
  group_by(id) %>%
  summarize(age_start = min(age_start),
            followup = max(TIME),
            nb_visits_left = sum(side == "left"),
            nb_visits_right = sum(side == "right"),
            nb_visits = n_distinct(TIME),
            nb_measurements = n(),
            visits_per_year = ifelse(followup == 0,
                                     NA,
                                     nb_visits / followup))
sum(summary_data$nb_visits_left != summary_data$nb_visits_right)#there are 112 people who weren't measure an equal amount of times on the left and right side
```

This report assesses hearing thresholds for a sample of `r nlevels(data$id)` healthy male volunteers. The subjects were `r round(mean(summary_data$age_start))` years old on average at the start of the study and are followed for an average of `r round(mean(summary_data$followup), digits = 2)` years. The hearing threshold is measured, on average, every `r round(1/mean(summary_data$visits_per_year, na.rm = T), digits = 2)` years, seperately for both ears.

Describe the data, and use graphical techniques to explore the mean structure, the
variance structure and the correlation structure. Summarize your conclusions. What are
the implications with respect to statistical modeling ?

# Methodology

# Results

# Discussion and conclusion

## further research

