---
documentclass: article
fontsize: 12pt
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2: 
    toc: false
    latex_engine: xelatex
    fig_caption: yes
    includes:
      in_header: preamble.sty
      before_body: titlepage.sty
bibliography: references.bib  
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven) # to read SAS files
library(kableExtra)
library(qwraps2)
library(rprojroot)
library(patchwork)
library(psych)
library(gtools)

knitr::opts_chunk$set(echo = TRUE)
```

```{r load, include=FALSE, message=FALSE, warning=FALSE}
trichotomization <- c(0, 20, 30, 100)
data <- read_sas(find_root_file("data/hearing500lr.sas7bdat", 
     criterion = has_file("LDA.Rproj"))) %>%
  mutate(side = as.factor(side),
         id = as.factor(id),
         age_measurement = age + TIME,
         age_discrete = cut(age,
                            breaks = c(0,30,50,70,100),
                            labels = c("<30", "30-50", "50-70",">70")),
         y_discrete = cut(y,
                            breaks = trichotomization,
                            labels = c("Normal", "Reduced", "Bad")),
         learning = 1*(TIME == 0))
```
# Data trichotomization

First, the data is trichotomized. The Centers for Disease Control and Prevention distinguished the following levels of hearing loss [@cdchearing], based on @clark1981:

- -10 - 25 dB: Normal hearing
- 26 - 40 dB: Mild hearing loss
- 41 - 55 dB: Moderate hearing loss
- 56 - 70 dB: Moderate / severe hearing loss
- 71 - 90 dB: Severe loss
- \geq 91 dB: Profound loss

It is quite common to dichotomize hearing loss into normal hearing ($\leq$25 dB) and hearing loss (>25 dB) [@garinis2017cumulative; @gallagher2019; @ju2022long] but thichotomization is less common. 

# Marginal model
Q2

# Random-effects model
Q2

## Empirical Bayes prediction
Q3

# Transition model
Q4

# Discussion


# Bibliography

